\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-27 1-intro.Rmd
chunk in changes

Spatio-temporal data involves variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube with three axes: spatial identifier, temporal identifier, and variable. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows four examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

@@

\textcolor{red}{start with "we need something to do exploratory data analysis for spatio-temporal data", then why it is hard to work with. This section needs to be abstract a bit further.}

None of these three formats in Figure \ref{fig:illu-input} is not particularly convenient to work with. In the single table format, to avoid plotting duplicated points of the same sites on the map, analysts always need to create a spatial variables table from the original single table. This can be a tedious step if needs to performed many times during the analysis. In another type of operation that filter the sites based on certain statistics, intermediate objects, like the per-station summary of the statistic and the filtered station ID vector, will need be created as standalone objects in the environment. The spirit of using pipe reduces the intermediate data objects that are unnecessarily created in the analysis while in this case, these only-used-once data objects can hardly be incorporated into a piping workflow in the analysis. In addition, if the data is organised as a separate spatial and temporal table as format 2 in Figure \ref{fig:illu-input}, the filter on site ID will need to performed twice in each table to keep them both up-to-date. With the array data structure, although it can easily store and wrangle large spatio-temporal data, the 3D or higher-D array can be less intuitive than a 2D tabular data structure for some analysts in data wrangling.

In the \proglang{R} community, different data objects have been proposed to work with spatio-temporal data. \pkg{spacetime} [@spacetime] is built on class \pkg{sp} [@sp] and \pkg{xts} [@xts] and both of which have been replaced by more recent spatial and temporal data objects. \pkg{spatstat} [@spatstat] implements a `ppp` class for analysing point pattern data; and more recent, \pkg{stars} [@stars] has implemented a spatio-temporal array for wrangling both vector and raster data. While these data objects are useful for its designed tasks, there has not yet been a data object that makes it easy to explore spatio-temporal data. Data objects like \pkg{sf} [@sf] and \pkg{tsibble} [@tsibble] are good examples of this in the spatial and temporal data analysis and hence there is a gap in software development for spatio-temporal data. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any data collected at fixed location across time.

<!-- While these implementations either store spatial and temporal variables all in a single table, hence duplicate the spatial variables at each temporal unit; or split them into two separate tables that analysts need to manually join the data constantly when doing the analysis. \textcolor{red}{There has been data structure proposed for spatial (\pkg{sf} by @sf) and temporal (\pkg{tsibble} by @tsibble) data in the R community, while less has been done for spatio-temporal data.}  -->

This paper describes a new tabular data object:  \pkg{cubble} for wrangling spatio-temporal data. \pkg{Cubble} overcomes the inconvenience in data wrangling mentioned above through wiring the separate spatial and temporal table into a single data object. With the new data object, analysts can spend less time organising spatial and temporal variables at different observational units and focus more on exploring the data itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

>>>>

Spatio-temporal data involves variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube with three axes: spatial identifier, temporal identifier, and variable. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows three examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

@@

While the data is enriched the information in both the spatial and temporal dimension, it also brings variables in two different observational units, which poses the problem of organising data for various operations in exploratory data analysis, visualisation, and modelling. The single table format is convenient for spatio-temporal operations, however, it makes duplicates of spatial variables and is not convenient for spatial operations. The two tables format allows separate spatial and temporal operations on each table while analysts need to work with two data objects simultaneously and dedicate efforts to make sure the sites in both tables are matched up. This data organising issue can also raise during the analysis, for example, a spatial summary of the temporal or spatio-temporal data naturally introduces an additional spatial table and makes the data in the separate table format. With the current tools, at each step, analysts need to decide on the format to operate on, perform the operation, and organise the result with existing data.

<!-- This would require analysts to constantly switch among different format at each step. All these difficulties are rooted in the two different observational units of spatial and temporal variables and this poses difficulties to organise spatio-temporal data in a way that operates both on a single spatial or temporal dimension and the combined spatio-temporal dimension. -->

<!-- None of these three formats in Figure \ref{fig:illu-input} is not particularly convenient to work with. In the single table format, to avoid plotting duplicated points of the same sites on the map, analysts always need to create a spatial variables table from the original single table. This can be a tedious step if needs to performed many times during the analysis. In another type of operation that filter the sites based on certain statistics, intermediate objects, like the per-station summary of the statistic and the filtered station ID vector, will need be created as standalone objects in the environment. The spirit of using pipe reduces the intermediate data objects that are unnecessarily created in the analysis while in this case, these only-used-once data objects can hardly be incorporated into a piping workflow in the analysis. In addition, if the data is organised as a separate spatial and temporal table as format 2 in Figure \ref{fig:illu-input}, the filter on site ID will need to performed twice in each table to keep them both up-to-date. With the array data structure, although it can easily store and wrangle large spatio-temporal data, the 3D or higher-D array can be less intuitive than a 2D tabular data structure for some analysts in data wrangling. -->

In the \proglang{R} community, different data objects have been proposed to work with spatio-temporal data when the specific task is known, while few have the capability to freely explore the data for exploratory data analysis. \pkg{spacetime} [@spacetime] is built on class \pkg{sp} [@sp] and \pkg{xts} [@xts] and both of which have been replaced by more recent spatial and temporal data objects. A more implementation, \pkg{stars} [@stars], uses a spatio-temporal array for wrangling both vector and raster data, but for raw data already in tabular format, need to convert the data into the array structure, also convert at vis. While there has not yet been a data object that makes it easy to explore spatio-temporal data that overcomes the difficulties of wrangling spatio-temporal data, this motivates developing a new tool that hopefully can solve this problem. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any data collected at fixed location across time. The requirement can also comes from spatial data analysts wanting to incorporate a temporal dimension in their data or vice versa.

This paper proposes an \proglang{R} package, \pkg{cubble}, for exploratory data analysis in spatio-temporal data. The proposed new data structure, also called \code{cubble}, overcomes the inconvenience in data wrangling mentioned above through wiring the separate spatial and temporal table into a single data object. With the new data object, analysts can spend less time organising spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

### 2022-02-27 main.Rmd
chunk in changes

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolas@langrene.com
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal data have two types of variables: spatial variables have fixed value across time for the same location while temporal variables need both the location and timestamp to identify its value. These variables can be either organised in two separate tables, each focusing on one aspect of the data, or a single table that combines the two types of variables together. Both formats are useful in exploratory data analysis and visualisation, however it can be tedious to constantly switch between the two formats during the analysis and spend time casting the data into the desireable shape. In this paper, a new data structure, \code{cubble}, is proposed to ease the transformation between the two formats in spatio-temporal data analysis. The new data structure organises spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while syncronised as a whole piece. More advanced topics discussed include handling data with hierarchical structure, matching data from multiple sources and using cubble in interactive graphics and spatio-temporal transformation. Examples are given using Australia climate weather station data, river level data, and climate reanalysis (ERA5) data to demonstrate the usage of the package.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

>>>>

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolas@langrene.com
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal data refers to measurements taken across space over time. In practise, spatial and temporal data many times need to be combined in order to form a spatio-temporal data set. However, these two types of data are collected at different observational units and this poses a problem on which format to use for different data wrangling tasks. Analysts can either organised the data in a single table format that combines all the variables together, or in two separate tables, each focusing on the spatial or temporal aspect of the data. Both formats are useful in exploratory data analysis, visualisation, and, modelling, however it can be labourious to constantly switch between the two formats during the analysis and write code to cast the data into the desireable shape. This paper presents a new \proglang{R} package, \pkg{cubble}, that provides a new data sturcture to ease the transformation between the two formats in spatio-temporal data. The new data structure organises spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while syncronised as a whole piece. \pkg{Cubble} also provides capability for handling data with hierarchical structure, matching data from multiple sources, constructing interactive graphics and performing spatio-temporal transformation. The paper will also demonstrate \pkg{cubble} with examples using Australia climate weather stations, river level data, and climate reanalysis (ERA5) data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

### 2022-02-27 main.tex
chunk in changes

\Abstract{
Spatio-temporal data have two types of variables: spatial variables have
fixed value across time for the same location while temporal variables
need both the location and timestamp to identify its value. These
variables can be either organised in two separate tables, each focusing
on one aspect of the data, or a single table that combines the two types
of variables together. Both formats are useful in exploratory data
analysis and visualisation, however it can be tedious to constantly
switch between the two formats during the analysis and spend time
casting the data into the desireable shape. In this paper, a new data
structure, \code{cubble}, is proposed to ease the transformation between
the two formats in spatio-temporal data analysis. The new data structure
organises spatial and temporal variables as two forms of a single data
object so that they can be wrangled separately while syncronised as a
whole piece. More advanced topics discussed include handling data with
hierarchical structure, matching data from multiple sources and using
cubble in interactive graphics and spatio-temporal transformation.
Examples are given using Australia climate weather station data, river
level data, and climate reanalysis (ERA5) data to demonstrate the usage
of the package.
}

@@

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time interval. This type of data can naturally
been considered as a data cube with three axes: spatial identifier,
temporal identifier, and variable. In this cubic formation, space only
occupies one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows four examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

\textcolor{red}{start with "we need something to do exploratory data analysis for spatio-temporal data", then why it is hard to work with. This section needs to be abstract a bit further.}

None of these three formats in Figure \ref{fig:illu-input} is not
particularly convenient to work with. In the single table format, to
avoid plotting duplicated points of the same sites on the map, analysts
always need to create a spatial variables table from the original single
table. This can be a tedious step if needs to performed many times
during the analysis. In another type of operation that filter the sites
based on certain statistics, intermediate objects, like the per-station
summary of the statistic and the filtered station ID vector, will need
be created as standalone objects in the environment. The spirit of using
pipe reduces the intermediate data objects that are unnecessarily
created in the analysis while in this case, these only-used-once data
objects can hardly be incorporated into a piping workflow in the
analysis. In addition, if the data is organised as a separate spatial
and temporal table as format 2 in Figure \ref{fig:illu-input}, the
filter on site ID will need to performed twice in each table to keep
them both up-to-date. With the array data structure, although it can
easily store and wrangle large spatio-temporal data, the 3D or higher-D
array can be less intuitive than a 2D tabular data structure for some
analysts in data wrangling.

In the \proglang{R} community, different data objects have been proposed
to work with spatio-temporal data. \pkg{spacetime} \citep{spacetime} is
built on class \pkg{sp} \citep{sp} and \pkg{xts} \citep{xts} and both of
which have been replaced by more recent spatial and temporal data
objects. \pkg{spatstat} \citep{spatstat} implements a \texttt{ppp} class
for analysing point pattern data; and more recent, \pkg{stars}
\citep{stars} has implemented a spatio-temporal array for wrangling both
vector and raster data. While these data objects are useful for its
designed tasks, there has not yet been a data object that makes it easy
to explore spatio-temporal data. Data objects like \pkg{sf} \citep{sf}
and \pkg{tsibble} \citep{tsibble} are good examples of this in the
spatial and temporal data analysis and hence there is a gap in software
development for spatio-temporal data. The requirement for such a tool is
important given the ubiquity of spatio-temporal vector data in the wild:
climate observations from weather stations, air quality variables from
air monitoring stations, or any data collected at fixed location across
time.

This paper describes a new tabular data object: \pkg{cubble} for
wrangling spatio-temporal data. \pkg{Cubble} overcomes the inconvenience
in data wrangling mentioned above through wiring the separate spatial
and temporal table into a single data object. With the new data object,
analysts can spend less time organising spatial and temporal variables
at different observational units and focus more on exploring the data
itself. The software is available from the Comprehensive R Archive
Network (CRAN) at {[}CRAN link{]}.

>>>>

\Abstract{
Spatio-temporal data refers to measurements taken across space over
time. In practise, spatial and temporal data many times need to be
combined in order to form a spatio-temporal data set. However, these two
types of data are collected at different observational units and this
poses a problem on which format to use for different data wrangling
tasks. Analysts can either organised the data in a single table format
that combines all the variables together, or in two separate tables,
each focusing on the spatial or temporal aspect of the data. Both
formats are useful in exploratory data analysis, visualisation, and,
modelling, however it can be labourious to constantly switch between the
two formats during the analysis and write code to cast the data into the
desireable shape. This paper presents a new \proglang{R} package,
\pkg{cubble}, that provides a new data sturcture to ease the
transformation between the two formats in spatio-temporal data. The new
data structure organises spatial and temporal variables as two forms of
a single data object so that they can be wrangled separately while
syncronised as a whole piece. \pkg{Cubble} also provides capability for
handling data with hierarchical structure, matching data from multiple
sources, constructing interactive graphics and performing
spatio-temporal transformation. The paper will also demonstrate
\pkg{cubble} with examples using Australia climate weather stations,
river level data, and climate reanalysis (ERA5) data.
}

@@

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time interval. This type of data can naturally
been considered as a data cube with three axes: spatial identifier,
temporal identifier, and variable. In this cubic formation, space only
occupies one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows three examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

While the data is enriched the information in both the spatial and
temporal dimension, it also brings variables in two different
observational units, which poses the problem of organising data for
various operations in exploratory data analysis, visualisation, and
modelling. The single table format is convenient for spatio-temporal
operations, however, it makes duplicates of spatial variables and is not
convenient for spatial operations. The two tables format allows separate
spatial and temporal operations on each table while analysts need to
work with two data objects simultaneously and dedicate efforts to make
sure the sites in both tables are matched up. This data organising issue
can also raise during the analysis, for example, a spatial summary of
the temporal or spatio-temporal data naturally introduces an additional
spatial table and makes the data in the separate table format. With the
current tools, at each step, analysts need to decide on the format to
operate on, perform the operation, and organise the result with existing
data.

In the \proglang{R} community, different data objects have been proposed
to work with spatio-temporal data when the specific task is known, while
few have the capability to freely explore the data for exploratory data
analysis. \pkg{spacetime} \citep{spacetime} is built on class \pkg{sp}
\citep{sp} and \pkg{xts} \citep{xts} and both of which have been
replaced by more recent spatial and temporal data objects. A more
implementation, \pkg{stars} \citep{stars}, uses a spatio-temporal array
for wrangling both vector and raster data, but for raw data already in
tabular format, need to convert the data into the array structure, also
convert at vis. While there has not yet been a data object that makes it
easy to explore spatio-temporal data that overcomes the difficulties of
wrangling spatio-temporal data, this motivates developing a new tool
that hopefully can solve this problem. The requirement for such a tool
is important given the ubiquity of spatio-temporal vector data in the
wild: climate observations from weather stations, air quality variables
from air monitoring stations, or any data collected at fixed location
across time. The requirement can also comes from spatial data analysts
wanting to incorporate a temporal dimension in their data or vice versa.

This paper proposes an \proglang{R} package, \pkg{cubble}, for
exploratory data analysis in spatio-temporal data. The proposed new data
structure, also called \code{cubble}, overcomes the inconvenience in
data wrangling mentioned above through wiring the separate spatial and
temporal table into a single data object. With the new data object,
analysts can spend less time organising spatial and temporal variables
and focus more on the exploratory data analysis itself. The software is
available from the Comprehensive R Archive Network (CRAN) at {[}CRAN
link{]}.
